### 2018-06-18 paper/paper.tex
Add initial, content-less version of paper

\documentclass[sigconf]{acmart}

\setcopyright{none}

\begin{document}

\title{Handwriting Recognition with SQL\\\hspace{6.88cm}\footnotesize and a tiny bit of web stuff}
\author{Noah Doersing}
\affiliation{}
\email{noah.doersing@student.uni-tuebingen.de}

\maketitle

\section*{Abstract}
This paper provides a sample of a \LaTeX\ document which conforms, somewhat loosely, to the formatting guidelines for ACM SIG Proceedings.\footnote{This is an abstract footnote}

\section{Introduction}

The main content goes here. testign

\section{Example/Demo?}

TODO really here? or something else?

\section{Implementation}

\subsection{Smoothing and Thinning}

test

\subsection{Curvature computation}

test

\subsection{Corner detection}

test

\subsection{Additional feature extraction}

test

\subsection{Mapping TODO}

test

\section{Evaluation}

test

\section{Related/Future work}

test

TODO conclusion, bibliography: essay, memo, memo on rand tablet, video of alan kay

\end{document}

### 2018-07-02 paper/paper.tex
Add notes

@@
 \title{Handwriting Recognition with SQL\\\hspace{6.88cm}\footnotesize and a tiny bit of web stuff}
-\author{Noah Doersing}
+\author{Noah Doersing TODO remove topmargin}
 \affiliation{}
@@
 This paper provides a sample of a \LaTeX\ document which conforms, somewhat loosely, to the formatting guidelines for ACM SIG Proceedings.\footnote{This is an abstract footnote}
+
+benjamin's idea worse because
+* 4-direction patterns in the first lookup table that map to a *single* character essentially skip the second lookup table (see `INSERT` in line 116 of the setup file).
+This could be somehow emulated using a query (if `COUNT(candidate\_character)` for the current pattern is 1, skip the feature-matching step), but that adds cruft to the query.
+* Deciding between 7 and 1 requires features F, while deciding between 7 and 3 requires looking at features G != F. Could also be patched over at the query level.
+* More verbose first lookup table, slimmer second lookup table.
+* Second lookup table does not anymore indicate which 4-direction pattern a given feature match belongs to, making debugging/extending harder.
+* Effectively requires ranking results based on closeness of match, e.g.
+number of features used (or VERY careful design of the second lookup table, which I didn't have time for before my presentation).
+* Conceptually further from original tree structure, which can be a bad or good thing.
 

### 2018-08-23 paper/paper.tex
Add notes
Add current state of paper

@@
 \documentclass[sigconf]{acmart}
+\setcopyright{none}
+\settopmatter{printfolios=true}
 
-\setcopyright{none}
+\usepackage{csquotes}
+\usepackage{fontspec}
+\setmonofont[Scale=0.88]{Iosevka}
+\usepackage{minted}
+\usepackage{subcaption}
+\usepackage{wasysym}
+\usepackage{pgfplots}
 
@@
 
-\title{Handwriting Recognition with SQL\\\hspace{6.88cm}\footnotesize and a tiny bit of web stuff}
-\author{Noah Doersing TODO remove topmargin}
+\title{\vspace{-0.4em}{\hspace{-6cm}\footnotesize \textsc{SQL is a Programming Language}}\\\vspace{-0.3em}Handwriting Recognition with SQL\\\vspace{-0.1em}\hspace{6.88cm}\footnotesize and a tiny bit of web stuff\vspace{-1em}}
+\author{Noah Doersing}
 \affiliation{}
@@
 \section*{Abstract}
-This paper provides a sample of a \LaTeX\ document which conforms, somewhat loosely, to the formatting guidelines for ACM SIG Proceedings.\footnote{This is an abstract footnote}
+A basic handwriting recognition algorithm conceived in the 60's has been implemented, 50 years later, on top of a modern relational database system.
+We discuss the origins of the system, introduce the algorithm and its SQL implementation, provide a web app that captures pen strokes and sends them off to the database for processing, evaluate the approach and provide further context.
 
-benjamin's idea worse because
+\section{Introduction}
+
+Throughout the history of computing, human-computer interaction has been an ever-evolving avenue for research.
+The results of computations, in the early days indicated by blinking light bulbs, have more recently been presented on screens of ever-increasing resolution, with different output formats such as voice being explored continuously.
+On the input side of the equation, punch cards have given way to command prompts fed by keyboards, which in turn have largely been superseded touch- or voice-based user interfaces.
+
+Both need to operate in lockstep: A fine meshing of input and output methods is required if an interface is to feel natural to users.
+Today's touch-based input devices, when compared to now-old-fashioned mouse input, exhibit clear usability upsides.
+Whenever the user wishes to manipulate a thing, she can simply touch it, instead of negotiating a mouse pointer towards it or entering keyboard shortcuts.
+
+Current as this may be, it is not fundamentally new.
+More than 50 years ago, researchers at RAND Corporation came up with a pen-based input paradigm intended, according to GUI pioneer Alan Kay, \enquote{for economists and other non-computer specialists [...] who said \enquote{Hey, none of us can type, can't you do something about that?}} \cite{kay} As part of this system, a basic handwriting recognition algorithm was developed.
+\cite{groner,schaedler}
+
+\textbf{Outline.} Before delving into its details – and how a simplified variant of it, only able to recognize a single-stroke letter at a time (see figure \ref{chars}), has been implemented as a SQL query – we will take a look at the RAND tablet and the GRAIL system it powers.
+After discussing the implementation, we will consider its performance and some related ideas.
+
+\begin{figure}[pb]
+  \centering
+  \includegraphics[width=7cm]{drawingguide}
+  \caption{How to draw the simplified characters.}
+  \label{chars}
+\end{figure}
+
+\section{GRAIL and the RAND tablet}
+
+\begin{figure}[tpb]
+  \centering
+  \includegraphics[width=7cm]{grailconsole}
+  \caption{A scan \cite{grail} of a printout of a photograph of a user interacting with GRAIL.}
+  \label{grailconsole}
+\end{figure}
+
+Developed by RAND in the 1960s, the Graphic Input Language (GRAIL) was a flowchart-based programming system.
+\cite{grail} As described by its creators, it enables a user to \enquote{draw flowchart symbols, edit and rearrange them on the display surface, and connect them appropriately to form a meaningful program.
+[The user can] execute the program while controlling its execution rate and the amount and content of information presented to him.
+The system interprets, in real-time, the [user's] hand-drawn figures, characters, and other [...] gestures to provide germane responses on the display surface.} \cite{grailang} Alan Kay, in a talk where he demonstrates GRAIL, summarizes it as \enquote{the world's first modeless system} and notes that it is \enquote{where Macintosh window control came from} \cite{kay}.
+
+The user operated the GRAIL system using a stylus connected to a tablet as shown in figure \ref{grailconsole}.
+Its \enquote{writing surface is a 10"x10" area with a resolution of 100 lines per inch in both x and y [...], allowing the user to \enquote{write} in a natural manner.
+[...] To maintain \enquote{naturalness} of the pen device, a pressure-sensitive switch in the tip of the stylus indicates \enquote{stroke} [...] to the computer} \cite{tablet}.\footnote{The authors of this memo describe the function of the tablet in great detail.
+Another memo takes a deep dive into the innards of GRAIL.
+\cite{details}}
+
+The original \enquote{IBM/360 assembly language program for online graphical-input character recognition using the RAND Tablet} \cite{source} is publicly available, but was not consulted when writing the SQL query – it is too low-level to be of much utility.
+
+\subsection{Web app}
+
+Testing the SQL query necessitates a simple simulation of the RAND tablet, which was given along with the assignment in the form of a D3.js-based web app that prints a list of coordinates once the user draws a character.
+
+In order to present the algorithm succinctly, this was extended with a server component that, with the help of some Python code, dynamically executes the handwriting recognition query on a Postgres instance and returns the result to the user.\footnote{This is given in \texttt{code/index.html}.
+The backend setup is a can of worms, so it's recommended to just use the frontend, which yields a JSON version of the pen stroke.}
+
+\begin{figure}[tpb]
+  \centering
+  \setminted{fontsize=\fontsize{6.04}{5.2}, linenos}
+  \begin{minted}{sql}
+WITH RECURSIVE
+tablet(pos, x, y) -- ... (conversion of JSON list to rows)
+smooth(pos, x, y) AS (
+  SELECT pos, x :: real, y :: real FROM tablet WHERE pos = 1
+    UNION ALL
+  SELECT t.pos,
+         (:smoothingfactor * s.x + (1.0 - :smoothingfactor) * t.x) :: real AS x,
+         (:smoothingfactor * s.y + (1.0 - :smoothingfactor) * t.y) :: real AS y
+  FROM   smooth s, tablet t
+  WHERE  t.pos = s.pos + 1
+),
+thin(pos, x, y) AS (
+  SELECT * FROM smooth WHERE pos = 1
+    UNION ALL
+  SELECT *
+  FROM   (
+    SELECT s.pos, s.x, s.y
+    FROM   thin t, smooth s
+    WHERE  s.pos > t.pos
+    AND    :thinningsize < |/ (s.x - t.x)^2 + (s.y - t.y)^2
+    ORDER BY s.pos
+    LIMIT 1
+  ) AS _
+),
+curve(pos, x, y, direction) AS (
+  SELECT pos, x, y,
+         COALESCE(degrees(-atan2( y - lag(y) OVER (ORDER BY pos),
+                                 -x + lag(x) OVER (ORDER BY pos))
+                         ) + 180, 90)
+  FROM   thin
+),
+cardinal(pos, direction) AS (
+  SELECT pos,
+         (enum_range(NULL :: cardinal_direction))[(direction / 90) :: int % 4 + 1]
+  FROM   curve
+),
+cardinal_change(pos, direction) AS (
+  SELECT pos, direction
+  FROM   (SELECT pos, direction,
+                 COALESCE(lag(direction, 2) OVER win <> lag(direction) OVER win,
+                          true)
+                 AND lag(direction) OVER win = direction
+          FROM cardinal
+          WINDOW win AS (ORDER BY pos)) AS _(pos, direction, is_new)
+  WHERE  is_new
+),
+corner(pos, x, y) AS (
+  SELECT pos, x, y
+  FROM   (SELECT pos, x, y, (
+                   angdiff(lag(direction, 2) OVER win, lag(direction) OVER win) < 22.5
+                   AND angdiff(lag(direction) OVER win, direction) > :cornerangle
+                   AND angdiff(direction, lead(direction) OVER win) < 22.5
+                 ) OR (  -- Immediate direction change OR one-segment turn.
+                   angdiff(lag(direction, 3) OVER win, lag(direction, 2) OVER win) < 22.5
+                   AND angdiff(lag(direction, 2) OVER win, direction) > :cornerangle
+                   AND angdiff(direction, lead(direction) OVER win) < 22.5
+                 ) AS is_corner
+          FROM   curve
+          WINDOW win AS (ORDER BY pos)) AS _(pos, x, y, is_corner)
+  WHERE  is_corner
+),
+aabb(xmin, xmax, ymin, ymax, aspect, width, height, centerx, centery) -- ...
+start_grid(n) AS (
+  SELECT gridpos(a.width, a.height, a.xmin, a.ymin, s.x, s.y)
+  FROM   smooth s, aabb a
+  ORDER BY s.pos
+  LIMIT 1
+),
+stop_grid(n) -- ... (analogous to start_grid)
+corner_grid(pos, n) -- ... (also similar)
+features(directions, start, stop, corners, width, height, aspect, center) AS (
+  SELECT (SELECT array_agg(direction ORDER BY pos)
+          FROM   cardinal_change),
+         (TABLE start_grid),
+         (TABLE stop_grid),
+         (SELECT COALESCE(array_agg(n ORDER BY pos), '{}')
+          FROM   corner_grid),
+         width, height, aspect, point(centerx, centery)
+  FROM   aabb
+),
+character(character) AS (
+  SELECT l.character
+  FROM   features f, lookup_candidates c, lookup_bestfit l
+  WHERE  f.directions[1:4] = c.first_four_directions
+  AND    c.candidate_characters = l.candidate_characters
+  AND    (l.start IS NULL          OR f.start = l.start)
+  AND    (l.stop IS NULL           OR f.stop = l.stop)
+  AND    (l.corners IS NULL        OR f.corners = l.corners)
+  AND    (l.aspect_range IS NULL   OR l.aspect_range @> f.aspect :: numeric)
+)
+TABLE character;
+  \end{minted}
+  \caption{Slightly condensed SQL implementation. Some custom functions have been omitted.}
+  \label{code}
+\end{figure}
+
+\section{Implementation}
+
+The web app outputs a JSON value of the form \texttt{[\{"x": 37, "y": 39\}, \{"x": 38, "y": 43, ...\}]}.
+It is passed to the Postgres\footnote{Postgres 9.4 or newer is required.} command-line client \texttt{psql} and converted to a table with an sequantial identifier column in CTE \texttt{tablet} (see figure \ref{code}).
+
+\subsection{Smoothing \& Thinning}
+
+\begin{figure}[pb]
+  \centering
+  \includegraphics[width=8.6cm]{smoothingthinning}
+  \caption{The {\color[HTML]{BE7777}\huge\textbullet} pen stroke drawn by the user, the corresponding {\color[HTML]{52B1C6}\huge\textbullet} quantized signal received by the computer, the {\color[HTML]{777777}\huge\textbullet} smoothed intermediate result and {\color[HTML]{74C974}\huge\textbullet} the points left after thinning.}
+  \label{smoothingthinning}
+\end{figure}
+
+The RAND tablet provides a stream (\enquote{the recognition scheme is notified of [the pen] position every 4 msec} \cite{groner}) of pen positions as integer $x,y$ coordinate pairs.
+
+As a preprocessing step, this digital representation pen stroke is smoothed to counteract quantization effects (\enquote{noise due to the discreteness of the pen location as measures by the tablet} \cite{groner}, which can negatively affect later steps of the algorithm) as shown in figure \ref{smoothingthinning}.
+Smoothing is performed by computing a weighted average of the most recently smoothed point (weight ~75\%; the first point of the stroke is assumed to be smooth already) and the next incoming point (~25\%).
+In SQL, this is accomplished by the recursive CTE \texttt{smooth} where the weight is stored in the \texttt{psql} variable \texttt{smoothingfactor}.
+
+The second stage of preprocessing drops points that are in close proximity to a previous point and thus contribute little or no additional information.
+This mainly serves to ease the computing effort required during the next stages.
+Groner uses a set of inequalities that compare $x$ and $y$ positions, effectively superimposing a square over each point and removing any other points that fall into this square.
+This is less \enquote{correct} than pruning based on the Euclidean distance as implemented in the SQL query (line 20), but computationally more efficient.
+The seemingly superfluous wrapping for the recursive part of the \texttt{thin} CTE in a \texttt{SELECT * FROM (...) AS \_} is required because Postgres does not permit \texttt{ORDER BY} directly in a recursive query.
+
+The degree to which these steps affect the raw data is controlled using the \texttt{psql} variables \texttt{smoothingfactor} and \texttt{thinningsize}.
+Choosing sensible values is important: As \texttt{smoothingfactor} approaches 1, corners disappear and the end of the stroke is pulled closer to the start.
+This effect is visible in figure \ref{smoothingthinning} where {\color[HTML]{BE7777}\huge\textbullet}{\color[HTML]{52B1C6}\huge\textbullet} extend further along the $x$ axis than {\color[HTML]{777777}\huge\textbullet}{\color[HTML]{74C974}\huge\textbullet}.
+Similarly, a large \texttt{thinningsize} can obscure corners or bends in the stroke.
+
+\subsection{Curvature}
+
+\enquote{Curvature is the most obvious characteristic which is independent of position and size, and yet which describes the [stroke]'s shape.
+[...] Four directions, used in conjunction with other features, provide sufficient description for recognition, yet result in [less effort] than do [more] directions.} \cite{groner} Only changes of these cardinal directions the pen stroke moves toward as each character is drawn are kept.
+
+Groner extracts this feature by comparing positional differences of point pairs using a set of inequalities.
+Because more fine-grained curvature information is helpful in detecting corners later on, the SQL implementation (see CTE \texttt{curve}) instead computes the precise angle between the line connecting each pair of points and the $x$ axis as shown in figure \ref{tan}.
+
+\begin{figure}[htpb]
+  \centering
+  \begin{tikzpicture}
+    \begin{axis}[
+      ticks=none,
+      axis lines = center,
+      axis line style={->},
+      ymin=-0.6, ymax=4.3,
+      xmin=-1, xmax=10,
+      axis equal,
+      xlabel=$x$, ylabel=$y$,
+      height=4.8cm,width=8cm]
+      \addplot[black, domain=0:8] {x*0.5};
+      \draw[black,fill=black] (axis cs:0,0) circle [radius=1] node[below right]{\tiny $p_1 = (x_1,y_1)$};
+      \draw[black,fill=black] (axis cs:8,4) circle [radius=1]  node[above left]{\tiny $p_2 = (x_2,y_2)$};
+      \draw (axis cs:2,0) arc[radius=1cm,start angle=0,end angle=31] node[left,yshift=-0.3cm,xshift=0.1cm]{$\alpha$} node[above,xshift=0.35cm,yshift=1.2cm]{\footnotesize$\tan(\alpha) = \frac{a}{b} = \frac{y_2-y_1}{x_2-x_1}$} node[below,xshift=0.43cm,yshift=1.35cm]{\footnotesize$\Rightarrow \alpha = \arctan\left(\frac{y_2-y_1}{x_2-x_1}\right)$};
+      \draw[dotted] (axis cs:8,4) -- node[right]{$a$} (axis cs:8,0);
+      \draw[dotted] (axis cs:0,-.02) -- node[below]{$b$} (axis cs:8,-.02);
+    \end{axis}
+  \end{tikzpicture}
+  \caption{Deriving the angle $\alpha$ between two points $p_1$ and $p_2$.}
+  \label{tan}
+\end{figure}
+
+In CTE \texttt{cardinal} it is then quantized to determine the cardinal directions, denoted by an \texttt{ENUM} type comprising the Unicode symbols $\blacktriangle\hspace{-0.15em}\blacktriangledown\hspace{-0.41em}\blacktriangleleft\blacktriangleright$.
+To accomplish this, the \texttt{enum\_range(anyenum)} function is used, which returns all values of the input enum type as an array.\footnote{See \url{https://www.postgresql.org/docs/8.3/static/functions-enum.html}.}
+
+\enquote{If the same direction occurs twice in succession, and is not the same as the last direction [...], then it is added to the [result], otherwise it is discarded.} \cite{groner} Thus the CTE \texttt{cardinal\_change} takes care of discarding sequential duplicates.
+
+TODO because window functions cannot be used in where, an awkward subquery todo "They are forbidden elsewhere, such as in \texttt{GROUP BY}, \texttt{HAVING} and \texttt{WHERE} clauses.
+This is because they logically execute after the processing of those clauses." \footnote{See \url{https://www.postgresql.org/docs/10/static/tutorial-window.html}.}
+
+\begin{figure}[pb]
+  \centering
+  \subcaptionbox{Corners help discern between similar charachters. They are determined based on the angles between four to five successive line segments.\label{corners}}{
+    \includegraphics[height=3.9cm]{corners}
+  }
+  \hspace{0.5cm}
+  \subcaptionbox{The absolute positions of corners, as well as start and end of a stroke are converted to grid positions.\label{grid}}{
+    \hspace{0.4cm}\includegraphics[height=3.9cm]{grid}\hspace{0.4cm}
+  }
+  \caption{Cardinal directions, corners and the grid.}
+  \label{cornergrid}
+\end{figure}
+
+\subsection{Corners}
+
+Cardinal directions are sufficient to discern between many characters, but in some cases – see figure \ref{corners} – other features such as the presence or absence of corners are required.
+\enquote{A corner is detected whenever the pen moves in the same ($\pm 1$) 16-direction for at least two segments [{\color[HTML]{21C800}$\blacktriangleright\blacktriangleright$}], changes direction by at least 90°, and then proceeds along the new direction ($\pm 1$) for at least two segments [{\color[HTML]{008AFF}$\blacktriangleright\blacktriangleright$}].
+The change in direction must take place either immediately or through a one-segment turn [{\color[HTML]{FFC000}$\blacktriangleright$}].} \cite{groner} (The triangles correspond to the lower half of figure \ref{corners}.)
+
+Instead of 16-directions (\textit{i.e.}, the division of 360° into 16 slices), the CTE \texttt{corner} of the SQL implementation uses absolute angles, making sure that the two line segments directly before a corner point within $\frac{360°}{16} = 22.5°$ of each other, the two next line segments also point within $22.5°$ of each other and the angle at the corner point is $>90°$.
+The window function \texttt{lag()} is used to access previous and next points – rather, the angles the associated line segments point towards – and a custom \texttt{angdiff()} function computes the delat $$\arctan\left(\frac{\sin(\alpha-\beta)}{\cos(\alpha-\beta)}\right)$$ between a pair of angles.\footnote{See \url{https://stackoverflow.com/a/2007279}.
+The formula $180 - \left|\left|\alpha - \beta\right| - 180\right|$ is equivalent and cheaper to compute, but the one given in the text elegantly combines sin, cos and (arc)tan.} The output of the CTE is a table of corner positions.
+
+\subsection{Additional features}
+
+In addition to cardinal directions and corners, Groner's algorithm determines a number of additional features: \enquote{the symbol's height and width [...], its aspect ratio [...], and its center relative to the tablet origin.} \cite{groner} They are of limited use to the simplified single-stroke alphanumeric characters supported by the SQL implementation; \textit{e.g.}, height and width aid in discerning parentheses from commata.
+
+Nevertheless, they are extracted in the CTE \texttt{aabb}, which trivially derives an axis-aligned bounding box around the character drawn by the user and returns two $x,y$ pairs defining it, along with width, height, center and aspect ratio.
+
+\textbf{Grid.} The algorithm then \enquote{divides the rectangular area defined by the symbol into a 4 x 4 grid.
+The starting (pen-down) and ending (pen-up) points, as well as the corner locations, are then each encoded as lying in one of these 16 areas, thereby locating them relative to the symbol.} \cite{groner} Figure \ref{grid} shows an example.
+
+The SQL implementation uses a custom \texttt{gridpos(width, height, xmin, ymin, x, y)} function which \enquote{desugars} to an arithmetic expression mapping \texttt{x} and \texttt{y}, given an AABB defined by minimum x, y, \texttt{width} and \texttt{height} to grid positions.\footnote{Global state would be convenient here to avoid passing the unchanging AABB on each call, but creating an extra temporary table for storing it seems like overkill.}
+
+Finally, the CTE \texttt{features} combines all extracted features into a single-row table.
+
+\subsection{Mapping Features to Characters}
+
+The
+
+TODO decision tree, not relational, most challenging portion of the work (rest smooth sailing)
+
+\textbf{Alternate approaches.} TODO benjamin's idea worse because
 * 4-direction patterns in the first lookup table that map to a *single* character essentially skip the second lookup table (see `INSERT` in line 116 of the setup file).
@@
 
-\section{Introduction}
-
-The main content goes here. testign
-
-\section{Example/Demo?}
-
-TODO really here? or something else?
-
-\section{Implementation}
-
-\subsection{Smoothing and Thinning}
-
-test
-
-\subsection{Curvature computation}
-
-test
-
-\subsection{Corner detection}
-
-test
-
-\subsection{Additional feature extraction}
-
-test
-
-\subsection{Mapping TODO}
-
-test
-
 \section{Evaluation}
 
-test
+Basing his analysis on the original IBM/360 assembly program running on a Model 40, Groner writes that the preprocessing and feature extraction stage \enquote{takes up about 40 percent of the available computing time in the limiting case [...].
+Smoothing and thinning requires about 20 percent of this analysis time, [drawing each incoming segment to the screen] 30 percent, and corner detection 16 percent.
+The remainder of the analysis time is for computing quantized directions, updating x and y extremes, and bookkeeping.} He goes on to outline potential improvements that could cut the computing time down to \enquote{only about 15 percent} \cite{groner}.
+TODO nothing about decision tree performance, probably quite fast
 
-\section{Related/Future work}
+TODO groner's user tests
 
-test
+Profiling the web version and SQL query is a bit more difficult TODO EXPLAIN analyze, compare to groner's breakdown, should roughly correspond
 
-TODO conclusion, bibliography: essay, memo, memo on rand tablet, video of alan kay
+\section{Related \& future work}
+
+\textbf{Related work.} TODO that moto? input thingy
+
+TODO dynamic time warping, see denis and http://www.hcii-lab.net/lianwen/Papers/Conference/%5BICDAR%202005%5D%20Hybrid%20recognition%20for%20one%20stroke%20style%20cursive%20handwriting%20characters.pdf
+
+TODO get the apple newton into this
+
+\textbf{Future work.} TODO improvements: multiple strokes, changes to make thing more suited to modern stuff (no smoothing/thinning due to higher (scan) resoltion of current tech), better representation of decision tree (perhaps more elegant on a graph db, but really no expert, heh)
+
+TODO some discussion on this approach, with multiple strokes support, being particularly suited for CJK input due to rigidly defined stroke order even though characters can look a bit complex, and indeed this has been done before https://www.rand.org/pubs/papers/P3568.html
+
+\section{Conclusion}
+
+TODO whatever, maybe drop
+
+\bibliographystyle{abbrv}
+\bibliography{paper.bib}
 

### 2018-08-24 paper/paper.tex
Add non-proofread variant of paper

@@
 \usepackage{pgfplots}
+\usepackage{enumitem}
+
+\urlstyle{tt}
 
@@
 
-\subsection{Web app}
+\subsection{Web App}
 
@@
 
-In CTE \texttt{cardinal} it is then quantized to determine the cardinal directions, denoted by an \texttt{ENUM} type comprising the Unicode symbols $\blacktriangle\hspace{-0.15em}\blacktriangledown\hspace{-0.41em}\blacktriangleleft\blacktriangleright$.
+In CTE \texttt{cardinal} it is then quantized to determine the cardinal directions, denoted by an \texttt{ENUM} type comprising the Unicode symbols $\blacktriangle\hspace{-0.15em}\blacktriangledown\hspace{-0.45em}\blacktriangleleft\blacktriangleright$.
 To accomplish this, the \texttt{enum\_range(anyenum)} function is used, which returns all values of the input enum type as an array.\footnote{See \url{https://www.postgresql.org/docs/8.3/static/functions-enum.html}.}
@@
 
-TODO because window functions cannot be used in where, an awkward subquery todo "They are forbidden elsewhere, such as in \texttt{GROUP BY}, \texttt{HAVING} and \texttt{WHERE} clauses.
+TODO because window functions cannot be used in where, an awkward subquery...
+"They are forbidden elsewhere, such as in \texttt{GROUP BY}, \texttt{HAVING} and \texttt{WHERE} clauses.
 This is because they logically execute after the processing of those clauses." \footnote{See \url{https://www.postgresql.org/docs/10/static/tutorial-window.html}.}
@@
 
-\subsection{Additional features}
+\subsection{Additional Features}
 
@@
 
-The
+Once the pen stroke has been sufficiently analyzed, Groner's algorithm \enquote{[looks] up the first four direction segments} in a table.
+For the letter B as shown in figure \ref{grid}, this table entry is $\blacktriangle\hspace{-0.26em}\blacktriangleright\hspace{-0.40em}\blacktriangledown\hspace{-0.40em}\blacktriangleleft$, which corresponds to 2, 3, 8, B, D, P, or R.
+\enquote{Further tests, based on this particular set of possibilities, result in a single identification.} \cite{groner}
 
-TODO decision tree, not relational, most challenging portion of the work (rest smooth sailing)
+In the original implementation, these tests take the form of a custom decision tree for each entry of the initial lookup table.
+A direct transliteration to SQL is difficult, requiring a rat's tail of nested \texttt{CASE} expressions or a similarly convoluted approach.
+However, it turned out that the simplified alphabet used in the context of this work suggests a more idiomatic two-stage approach that, while less flexible, delivers acceptable results:
 
-\textbf{Alternate approaches.} TODO benjamin's idea worse because
-* 4-direction patterns in the first lookup table that map to a *single* character essentially skip the second lookup table (see `INSERT` in line 116 of the setup file).
+\begin{enumerate}[nolistsep]
+    \item An array comprising the first four cardinal directions is mapped to an array of possible characters using lookup table $T_1$.
+    \item Another lookup table $T_2$ is consulted that maps the potential characters and specific feature values – some of which can be marked as \enquote{don't care} by setting them to \texttt{NULL} (hence the \texttt{IS NULL} checks in lines 86 through 89) – to a concrete result character.
+For characters which are wholly determined by the first lookup, all features are set to \texttt{NULL}.
+\end{enumerate}
+
+Finding a satisfactory representation for this decision mechanism was one of the trickier aspects of this work: A reasonable balance between flexibility and maintainability needed to be struck while ideally still resembling Groner's ideas.
+
+\textbf{Alternate approach.} The two-stage approach implemented in CTE \texttt{character} is not ideal: The array-to-array mapping of $T_1$ does not conform to 1NF.
+An alternate representation\footnote{Its implementation can be found in the \texttt{morerelational} branch of the Git repository.} was explored where the rows of $T_1$ were \enquote{expanded} to map to single characters from direction arrays (which can occur multiple times in the table) and $T_2$ is modified accordingly by chopping off the array column.
+This, however, complicates CTE \texttt{character} and complicates the ability to discern some pairs of characters: Deciding whether a stroke resembles 7 or 1 requires different features than the choice between 7 and 3, say.
+This difficulty could be patched over on the query level, however.
+
+%TODO benjamin's idea worse because
+%* 4-direction patterns in the first lookup table that map to a *single* character essentially skip the second lookup table (see `INSERT` in line 116 of the setup file).
 This could be somehow emulated using a query (if `COUNT(candidate\_character)` for the current pattern is 1, skip the feature-matching step), but that adds cruft to the query.
-* Deciding between 7 and 1 requires features F, while deciding between 7 and 3 requires looking at features G != F. Could also be patched over at the query level.
-* More verbose first lookup table, slimmer second lookup table.
-* Second lookup table does not anymore indicate which 4-direction pattern a given feature match belongs to, making debugging/extending harder.
-* Effectively requires ranking results based on closeness of match, e.g.
+%* Deciding between 7 and 1 requires features F, while deciding between 7 and 3 requires looking at features G != F. Could also be patched over at the query level.
+%* More verbose first lookup table, slimmer second lookup table.
+%* Second lookup table does not anymore indicate which 4-direction pattern a given feature match belongs to, making debugging/extending harder.
+%* Effectively requires ranking results based on closeness of match, e.g.
 number of features used (or VERY careful design of the second lookup table, which I didn't have time for before my presentation).
-* Conceptually further from original tree structure, which can be a bad or good thing.
+%* Conceptually further from original tree structure, which can be a bad or good thing.
 
@@
 Smoothing and thinning requires about 20 percent of this analysis time, [drawing each incoming segment to the screen] 30 percent, and corner detection 16 percent.
-The remainder of the analysis time is for computing quantized directions, updating x and y extremes, and bookkeeping.} He goes on to outline potential improvements that could cut the computing time down to \enquote{only about 15 percent} \cite{groner}.
-TODO nothing about decision tree performance, probably quite fast
+The remainder of the analysis time is for computing quantized directions, updating x and y extremes, and bookkeeping.} He goes on to outline potential improvements that could cut the computing time down to \enquote{only about 15 percent} \cite{groner} but does not mention the cost of walking the decision tree – supposedly it's minimal.
+Groner also presents the results of user testing.
 
-TODO groner's user tests
-
-Profiling the web version and SQL query is a bit more difficult TODO EXPLAIN analyze, compare to groner's breakdown, should roughly correspond
+The SQL implementation takes $10-15$ ms to process a typical pen stroke generated by the web app.
+Taking a look at the \texttt{EXPLAIN ANALYZE} output of the query shown in figure \ref{code}, none of the CTEs take an unexpectedly high amount of \texttt{actual time} – the computing effort seems to be relatively evenly spread, without any major departures from Groner's observations.
 
@@
 
-\textbf{Related work.} TODO that moto? input thingy
+\textbf{Related work.} GRAIL is a successor to Sketchpad, one of the first systems that \enquote{[make] it possible for a man and a computer to converse rapidly through the medium of line drawings} but relied on a keyboard for text input.
+\cite{sketchpad}
 
-TODO dynamic time warping, see denis and http://www.hcii-lab.net/lianwen/Papers/Conference/%5BICDAR%202005%5D%20Hybrid%20recognition%20for%20one%20stroke%20style%20cursive%20handwriting%20characters.pdf
+Because Chinese and Korean characters are drawn based on a rigidly defined stroke order, algorithms like the one presented here are well-suited for these writing systems.
+Indeed, Groner and Heafner worked on \enquote{classification of handprinted Chinese characters as a translation aid} \cite{chinese} in the same timeframe.
 
-TODO get the apple newton into this
+Palm later developed Graffiti, a text input system based on single-stroke characters and commands.
+Among other devices, it was deployed on Apple's Newton and Windows Mobile.\footnote{See \url{https://en.wikipedia.org/wiki/Graffiti_(Palm_OS)}.} The SQL implementation could conceivably be modified to recognize Graffiti's gestures by rewriting the lookup tables.
 
-\textbf{Future work.} TODO improvements: multiple strokes, changes to make thing more suited to modern stuff (no smoothing/thinning due to higher (scan) resoltion of current tech), better representation of decision tree (perhaps more elegant on a graph db, but really no expert, heh)
+Dynamic Time Warping may be used for character recognition, given an effective way of converting stroke information to time series.\footnote{Ask the guy in B312 about this.}
 
-TODO some discussion on this approach, with multiple strokes support, being particularly suited for CJK input due to rigidly defined stroke order even though characters can look a bit complex, and indeed this has been done before https://www.rand.org/pubs/papers/P3568.html
+\textbf{Future work.} Recognizing characters composed of multiple strokes would bring the SQL implementation to basic feature parity with the original system.
 
-\section{Conclusion}
+Groner's decision tree could be represented more effectively – building the character recognition scheme on top of a graph database might lend itself to an implementation that's more true to the original.
 
-TODO whatever, maybe drop
+Because modern tablets feature resolutions significantly higher than the RAND tablet, the smoothing step is less crucial now than it was when the RAND tablet was used.
+Similarly, the time savings brought about by thinning are barely worth the computing effort on modern, faster systems.
 

### 2018-08-24 paper/paper.tex
Add final version of paper

@@
 
-\usepackage{csquotes}
 \usepackage{fontspec}
 \setmonofont[Scale=0.88]{Iosevka}
+\newfontfamily\emoji{DejaVu Sans}
+
+\usepackage{csquotes}
 \usepackage{minted}
@@
 
-\title{\vspace{-0.4em}{\hspace{-6cm}\footnotesize \textsc{SQL is a Programming Language}}\\\vspace{-0.3em}Handwriting Recognition with SQL\\\vspace{-0.1em}\hspace{6.88cm}\footnotesize and a tiny bit of web stuff\vspace{-1em}}
+\title{\vspace{-0.5em}{\hspace{-5.5cm}\small \textsc{SQL is a Programming Language}}\\\vspace{-0.3em}Handwriting Recognition with SQL\\\vspace{-0.1em}\hspace{6.88cm}\footnotesize and a tiny bit of web stuff\vspace{-1em}}
 \author{Noah Doersing}
@@
 Throughout the history of computing, human-computer interaction has been an ever-evolving avenue for research.
-The results of computations, in the early days indicated by blinking light bulbs, have more recently been presented on screens of ever-increasing resolution, with different output formats such as voice being explored continuously.
-On the input side of the equation, punch cards have given way to command prompts fed by keyboards, which in turn have largely been superseded touch- or voice-based user interfaces.
+The results of computations, in the early days indicated by blinking light bulbs, have more recently been presented on screens of ever-increasing resolution, with different output formats being explored continuously.
+On the input side of the equation, punch cards have given way to command prompts fed by keyboards, which in turn have largely been superseded by touch- or voice-based user interfaces.
 
-Both need to operate in lockstep: A fine meshing of input and output methods is required if an interface is to feel natural to users.
-Today's touch-based input devices, when compared to now-old-fashioned mouse input, exhibit clear usability upsides.
-Whenever the user wishes to manipulate a thing, she can simply touch it, instead of negotiating a mouse pointer towards it or entering keyboard shortcuts.
+Both need to operate in lockstep: The meshing of input and output methods must be carefully considered if an interface is to feel natural to its users.
+Today's touch-based input devices, when compared to now-old-fashioned mouse input, exhibit clear usability upsides – when the user wishes to manipulate a thing, she can simply touch it, instead of negotiating a mouse pointer towards it or entering keyboard shortcuts.
 
-Current as this may be, it is not fundamentally new.
+\textit{Current} as this may be, it is not fundamentally \textit{new}.
 More than 50 years ago, researchers at RAND Corporation came up with a pen-based input paradigm intended, according to GUI pioneer Alan Kay, \enquote{for economists and other non-computer specialists [...] who said \enquote{Hey, none of us can type, can't you do something about that?}} \cite{kay} As part of this system, a basic handwriting recognition algorithm was developed.
@@
 
-Developed by RAND in the 1960s, the Graphic Input Language (GRAIL) was a flowchart-based programming system.
-\cite{grail} As described by its creators, it enables a user to \enquote{draw flowchart symbols, edit and rearrange them on the display surface, and connect them appropriately to form a meaningful program.
+Developed at RAND in the 1960s, the Graphic Input Language (GRAIL) was a flowchart-based programming system \cite{kay,grailang,grail}.
+As described by its creators, it enables a user to \enquote{draw flowchart symbols, edit and rearrange them on the display surface, and connect them appropriately to form a meaningful program.
 [The user can] execute the program while controlling its execution rate and the amount and content of information presented to him.
-The system interprets, in real-time, the [user's] hand-drawn figures, characters, and other [...] gestures to provide germane responses on the display surface.} \cite{grailang} Alan Kay, in a talk where he demonstrates GRAIL, summarizes it as \enquote{the world's first modeless system} and notes that it is \enquote{where Macintosh window control came from} \cite{kay}.
+The system interprets, in real-time, the [user's] hand-drawn figures, characters, and other [...] gestures to provide germane responses on the display surface.} \cite{grailang} Alan Kay, in a talk where he demonstrates GRAIL, summarizes it as \enquote{the world's first modeless system}
+%and notes that it is \enquote{where Macintosh window control came from}
+\cite{kay}.
 
@@
 The web app outputs a JSON value of the form \texttt{[\{"x": 37, "y": 39\}, \{"x": 38, "y": 43, ...\}]}.
-It is passed to the Postgres\footnote{Postgres 9.4 or newer is required.} command-line client \texttt{psql} and converted to a table with an sequantial identifier column in CTE \texttt{tablet} (see figure \ref{code}).
+It is passed to the Postgres\footnote{Postgres 9.4 or newer is required.} command-line client \texttt{psql} and converted to a table with a sequential identifier column in CTE \texttt{tablet} (see figure \ref{code}).
 
@@
 
-The RAND tablet provides a stream (\enquote{the recognition scheme is notified of [the pen] position every 4 msec} \cite{groner}) of pen positions as integer $x,y$ coordinate pairs.
+The RAND tablet provides a stream (\enquote{the recognition scheme is notified of [the pen] position every 4 msec} \cite{groner}) of pen positions as integer-valued $x,y$ coordinate pairs.
 
-As a preprocessing step, this digital representation pen stroke is smoothed to counteract quantization effects (\enquote{noise due to the discreteness of the pen location as measures by the tablet} \cite{groner}, which can negatively affect later steps of the algorithm) as shown in figure \ref{smoothingthinning}.
-Smoothing is performed by computing a weighted average of the most recently smoothed point (weight ~75\%; the first point of the stroke is assumed to be smooth already) and the next incoming point (~25\%).
-In SQL, this is accomplished by the recursive CTE \texttt{smooth} where the weight is stored in the \texttt{psql} variable \texttt{smoothingfactor}.
+As a preprocessing step, this digital representation of the pen stroke is smoothed to counteract quantization effects (\enquote{noise due to the discreteness of the pen location as measured by the tablet} \cite{groner}, which can negatively affect the latter steps of the algorithm) as shown in figure \ref{smoothingthinning}.
+Smoothing is performed by computing a weighted average of the most recently smoothed point (weight ~75\%; the first point of the stroke is accepted as is) and the next incoming point (~25\%).
+In SQL, this is accomplished by the recursive CTE \texttt{smooth} where the weight is read from the \texttt{psql} variable \texttt{smoothingfactor}.
 
-The second stage of preprocessing drops points that are in close proximity to a previous point and thus contribute little or no additional information.
-This mainly serves to ease the computing effort required during the next stages.
-Groner uses a set of inequalities that compare $x$ and $y$ positions, effectively superimposing a square over each point and removing any other points that fall into this square.
+The second stage of preprocessing, thinning, drops points that are in close proximity to the most recently accepted point and thus contribute little or no additional information.
+This mainly serves to ease the computing effort required during the next steps.
+Groner uses a set of inequalities that compare $x$ and $y$ positions, effectively superimposing a square over each thinned point (again, the first point is accepted as is) and removing any other points that fall into this square.
 This is less \enquote{correct} than pruning based on the Euclidean distance as implemented in the SQL query (line 20), but computationally more efficient.
-The seemingly superfluous wrapping for the recursive part of the \texttt{thin} CTE in a \texttt{SELECT * FROM (...) AS \_} is required because Postgres does not permit \texttt{ORDER BY} directly in a recursive query.
+The seemingly superfluous wrapping of the recursive part of the \texttt{thin} CTE in a \texttt{SELECT * FROM (...) AS \_} is required because Postgres does not permit \texttt{ORDER BY} directly in a recursive query.
 
 The degree to which these steps affect the raw data is controlled using the \texttt{psql} variables \texttt{smoothingfactor} and \texttt{thinningsize}.
-Choosing sensible values is important: As \texttt{smoothingfactor} approaches 1, corners disappear and the end of the stroke is pulled closer to the start.
+Choosing sensible values is crucial: As \texttt{smoothingfactor} approaches 1, corners disappear and the end of the stroke is pulled closer to the start.
 This effect is visible in figure \ref{smoothingthinning} where {\color[HTML]{BE7777}\huge\textbullet}{\color[HTML]{52B1C6}\huge\textbullet} extend further along the $x$ axis than {\color[HTML]{777777}\huge\textbullet}{\color[HTML]{74C974}\huge\textbullet}.
-Similarly, a large \texttt{thinningsize} can obscure corners or bends in the stroke.
+Similarly, a large \texttt{thinningsize} can obscure corners or bends, which character recognition hinges on.
 
@@
 \enquote{Curvature is the most obvious characteristic which is independent of position and size, and yet which describes the [stroke]'s shape.
-[...] Four directions, used in conjunction with other features, provide sufficient description for recognition, yet result in [less effort] than do [more] directions.} \cite{groner} Only changes of these cardinal directions the pen stroke moves toward as each character is drawn are kept.
+[...] Four directions, used in conjunction with other features, provide sufficient description for recognition, yet result in [less effort] than do [more] directions.} \cite{groner} The algorithm only keeps \textit{changes} of these cardinal directions the pen stroke moves toward as each character is drawn.
 
 Groner extracts this feature by comparing positional differences of point pairs using a set of inequalities.
-Because more fine-grained curvature information is helpful in detecting corners later on, the SQL implementation (see CTE \texttt{curve}) instead computes the precise angle between the line connecting each pair of points and the $x$ axis as shown in figure \ref{tan}.
+Because more fine-grained curvature information is helpful in detecting corners later on, the SQL implementation (see CTE \texttt{curve}) instead computes the precise angle between the imaginary line segment connecting each pair of points and the $x$ axis as shown in figure \ref{tan}.
 
@@
 
-In CTE \texttt{cardinal} it is then quantized to determine the cardinal directions, denoted by an \texttt{ENUM} type comprising the Unicode symbols $\blacktriangle\hspace{-0.15em}\blacktriangledown\hspace{-0.45em}\blacktriangleleft\blacktriangleright$.
-To accomplish this, the \texttt{enum\_range(anyenum)} function is used, which returns all values of the input enum type as an array.\footnote{See \url{https://www.postgresql.org/docs/8.3/static/functions-enum.html}.}
+In CTE \texttt{cardinal} $\alpha$ is then quantized to determine the cardinal directions, denoted by an \texttt{ENUM} type containing the Unicode symbols $\blacktriangle\hspace{-0.15em}\blacktriangledown\hspace{-0.45em}\blacktriangleleft\blacktriangleright$.
+To accomplish this, the \texttt{enum\_range(anyenum)} function, which returns all values of the input enum type as an array, is used.\footnote{See \url{https://www.postgresql.org/docs/8.3/static/functions-enum.html}.}
 
-\enquote{If the same direction occurs twice in succession, and is not the same as the last direction [...], then it is added to the [result], otherwise it is discarded.} \cite{groner} Thus the CTE \texttt{cardinal\_change} takes care of discarding sequential duplicates.
-
-TODO because window functions cannot be used in where, an awkward subquery...
-"They are forbidden elsewhere, such as in \texttt{GROUP BY}, \texttt{HAVING} and \texttt{WHERE} clauses.
-This is because they logically execute after the processing of those clauses." \footnote{See \url{https://www.postgresql.org/docs/10/static/tutorial-window.html}.}
+As mentioned above, \enquote{[i]f the same direction occurs twice in succession, and is not the same as the last direction [...], then it is added to the [result], otherwise it is discarded} \cite{groner}, which the CTE \texttt{cardinal\_change} takes care of.
+Because window functions cannot\footnote{\enquote{This is because they logically execute after the processing of [this clause]}, see \url{https://www.postgresql.org/docs/10/static/tutorial-window.html}.} be used in \texttt{WHERE} predicates, an somewhat awkward subquery builds a boolean column, based on which the rows are then filtered (lines 39 to 45).
 
@@
   \centering
-  \subcaptionbox{Corners help discern between similar charachters. They are determined based on the angles between four to five successive line segments.\label{corners}}{
+  \subcaptionbox{Corners help discern between similar characters. They are determined based on the angles between four to five successive line segments.\label{corners}}{
     \includegraphics[height=3.9cm]{corners}
@@
 
-Cardinal directions are sufficient to discern between many characters, but in some cases – see figure \ref{corners} – other features such as the presence or absence of corners are required.
+Cardinal directions are sufficient to discern between many characters, but in some cases, \textit{e.g.}, G and 6 as demonstrated in figure \ref{corners}, other features such as the presence or absence of corners are required.
 \enquote{A corner is detected whenever the pen moves in the same ($\pm 1$) 16-direction for at least two segments [{\color[HTML]{21C800}$\blacktriangleright\blacktriangleright$}], changes direction by at least 90°, and then proceeds along the new direction ($\pm 1$) for at least two segments [{\color[HTML]{008AFF}$\blacktriangleright\blacktriangleright$}].
@@
 
-Instead of 16-directions (\textit{i.e.}, the division of 360° into 16 slices), the CTE \texttt{corner} of the SQL implementation uses absolute angles, making sure that the two line segments directly before a corner point within $\frac{360°}{16} = 22.5°$ of each other, the two next line segments also point within $22.5°$ of each other and the angle at the corner point is $>90°$.
-The window function \texttt{lag()} is used to access previous and next points – rather, the angles the associated line segments point towards – and a custom \texttt{angdiff()} function computes the delat $$\arctan\left(\frac{\sin(\alpha-\beta)}{\cos(\alpha-\beta)}\right)$$ between a pair of angles.\footnote{See \url{https://stackoverflow.com/a/2007279}.
+Instead of 16-directions (\textit{i.e.}, the division of 360° into 16 slices), the CTE \texttt{corner} of the SQL implementation uses absolute angles, making sure that the angles corresponding to the two line segments directly before a corner are within $\frac{360°}{16} = 22.5°$ of each other, the angle at the corner point is $>90°$ and the two subsequent angles again differ by no more than $22.5°$.
+The window function \texttt{lag()} is used to access these previous and next angles and a custom \texttt{angdiff()} function computes the delta $$\arctan\left(\frac{\sin(\alpha-\beta)}{\cos(\alpha-\beta)}\right)$$ between each pair of angles.\footnote{See \url{https://stackoverflow.com/a/2007279}.
 The formula $180 - \left|\left|\alpha - \beta\right| - 180\right|$ is equivalent and cheaper to compute, but the one given in the text elegantly combines sin, cos and (arc)tan.} The output of the CTE is a table of corner positions.
@@
 
-In addition to cardinal directions and corners, Groner's algorithm determines a number of additional features: \enquote{the symbol's height and width [...], its aspect ratio [...], and its center relative to the tablet origin.} \cite{groner} They are of limited use to the simplified single-stroke alphanumeric characters supported by the SQL implementation; \textit{e.g.}, height and width aid in discerning parentheses from commata.
+In addition to cardinal directions and corners, Groner's algorithm determines a number of additional features: \enquote{the symbol's height and width [...], its aspect ratio [...], and its center relative to the tablet origin.} \cite{groner} They are of limited use to the simplified single-stroke alphanumeric characters recognized by the SQL implementation; \textit{e.g.}, height and width help discern parentheses from commata, neither of which are supported.
 
-Nevertheless, they are extracted in the CTE \texttt{aabb}, which trivially derives an axis-aligned bounding box around the character drawn by the user and returns two $x,y$ pairs defining it, along with width, height, center and aspect ratio.
+Nevertheless, they are extracted in the CTE \texttt{aabb}, which trivially derives an axis-aligned bounding box around the character drawn by the user and returns two $x,y$ coordinates defining it, along with width, height, center and aspect ratio.
 
@@
 
-The SQL implementation uses a custom \texttt{gridpos(width, height, xmin, ymin, x, y)} function which \enquote{desugars} to an arithmetic expression mapping \texttt{x} and \texttt{y}, given an AABB defined by minimum x, y, \texttt{width} and \texttt{height} to grid positions.\footnote{Global state would be convenient here to avoid passing the unchanging AABB on each call, but creating an extra temporary table for storing it seems like overkill.}
+The SQL implementation uses a custom \texttt{gridpos(width, height, xmin, ymin, x, y)} function which \enquote{desugars} to an arithmetic expression mapping \texttt{x} and \texttt{y}, given an AABB defined by minimum x, y, width and height to one of the 16 grid positions.\footnote{Global state would be convenient here to avoid passing the unchanging AABB on each call, but creating an extra temporary table for storing it seems like overkill.}
 
-Finally, the CTE \texttt{features} combines all extracted features into a single-row table.
+Finally, the CTE \texttt{features} combines all extracted features into a handy single-row table.
 
@@
 A direct transliteration to SQL is difficult, requiring a rat's tail of nested \texttt{CASE} expressions or a similarly convoluted approach.
-However, it turned out that the simplified alphabet used in the context of this work suggests a more idiomatic two-stage approach that, while less flexible, delivers acceptable results:
+However, the simplified alphabet used in the context of this work suggested a more idiomatic two-stage approach that, while less flexible, delivers acceptable results:
 
@@
     \item An array comprising the first four cardinal directions is mapped to an array of possible characters using lookup table $T_1$.
-    \item Another lookup table $T_2$ is consulted that maps the potential characters and specific feature values – some of which can be marked as \enquote{don't care} by setting them to \texttt{NULL} (hence the \texttt{IS NULL} checks in lines 86 through 89) – to a concrete result character.
+    \item Another lookup table $T_2$ is consulted, mapping the potential characters and specific feature values – some of which can be marked as \enquote{don't care} by setting them to \texttt{NULL}, hence the \texttt{IS NULL} checks in lines 86 through 89 – to a concrete result character.
 For characters which are wholly determined by the first lookup, all features are set to \texttt{NULL}.
@@
 \textbf{Alternate approach.} The two-stage approach implemented in CTE \texttt{character} is not ideal: The array-to-array mapping of $T_1$ does not conform to 1NF.
-An alternate representation\footnote{Its implementation can be found in the \texttt{morerelational} branch of the Git repository.} was explored where the rows of $T_1$ were \enquote{expanded} to map to single characters from direction arrays (which can occur multiple times in the table) and $T_2$ is modified accordingly by chopping off the array column.
-This, however, complicates CTE \texttt{character} and complicates the ability to discern some pairs of characters: Deciding whether a stroke resembles 7 or 1 requires different features than the choice between 7 and 3, say.
-This difficulty could be patched over on the query level, however.
+An alternate representation\footnote{Its implementation can be found in the \texttt{morerelational} branch of the Git repository.} was explored where the rows of $T_1$ were \enquote{expanded} to map to single characters from direction arrays (which can now occur multiple times in the table) and $T_2$ is modified accordingly by chopping off the array column.
+This, however, complicates the CTE \texttt{character} and complicates future expansion \textit{and} discerning some pairs of characters: Deciding whether a stroke resembles 7 or 1 requires different features than the choice between 7 and 3, say.
+This difficulty, however, could be patched over on the query level.
 
@@
 The SQL implementation takes $10-15$ ms to process a typical pen stroke generated by the web app.
-Taking a look at the \texttt{EXPLAIN ANALYZE} output of the query shown in figure \ref{code}, none of the CTEs take an unexpectedly high amount of \texttt{actual time} – the computing effort seems to be relatively evenly spread, without any major departures from Groner's observations.
+Taking a look at the output of \texttt{EXPLAIN ANALYZE}, none of the CTEs take an unexpectedly high amount of \texttt{actual time}.
+Thus, the computing effort seems to be relatively evenly spread, without any major departures from Groner's observations.
 
@@
 
-Because Chinese and Korean characters are drawn based on a rigidly defined stroke order, algorithms like the one presented here are well-suited for these writing systems.
+Because Chinese and Korean characters are drawn based on a rigidly defined stroke order, algorithms like the one presented are well-suited for these writing systems.
 Indeed, Groner and Heafner worked on \enquote{classification of handprinted Chinese characters as a translation aid} \cite{chinese} in the same timeframe.
 
-Palm later developed Graffiti, a text input system based on single-stroke characters and commands.
+Palm, Inc.
+later developed Graffiti, a text input system based on single-stroke characters and commands.
 Among other devices, it was deployed on Apple's Newton and Windows Mobile.\footnote{See \url{https://en.wikipedia.org/wiki/Graffiti_(Palm_OS)}.} The SQL implementation could conceivably be modified to recognize Graffiti's gestures by rewriting the lookup tables.
 
-Dynamic Time Warping may be used for character recognition, given an effective way of converting stroke information to time series.\footnote{Ask the guy in B312 about this.}
+DTW may be used for handwriting recognition, contingent on an effective way of converting pen strokes to time series.\footnote{Ask the guy in B312 about this. {\emoji 😉}}
+
+Jack Schaedler has build an excellent \textit{explorable explanation} \cite{schaedler} showcasing GRAIL's handwriting recognition scheme, which many visualizations used in this work are based on.
 
@@
 
-Groner's decision tree could be represented more effectively – building the character recognition scheme on top of a graph database might lend itself to an implementation that's more true to the original.
+Groner's decision tree could be represented more effectively – building the character recognition scheme on top of a graph database \textit{might} lend itself to an implementation that's more true to the original.
 
-Because modern tablets feature resolutions significantly higher than the RAND tablet, the smoothing step is less crucial now than it was when the RAND tablet was used.
-Similarly, the time savings brought about by thinning are barely worth the computing effort on modern, faster systems.
+Because modern tablets feature resolutions significantly higher than the RAND tablet, the smoothing step is less crucial now than it was back then.
+Similarly, the time savings brought about by thinning barely make a dent on modern systems.
 
